---
title: "Coursework 1: MA22019, Introduction to Data Science"
author: "Ewan Brett"
format: gfm
---

<!--- DO NOT DELETE THIS LINE --->
<!--- METADATA ANCHOR: MA22019-CW1-2026 --->

Please read the `coursework_01_instructions.pdf` for the full context, data dictionary, and detailed task requirements before starting.



```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(tidytext)
library(knitr)
library(patchwork)
```

## Part A: Data Wrangling & EDA 

### Task 1: Data Joining
<!--- DO NOT DELETE THIS LINE - TASK 1 ANCHOR --->

```{r}
#| label: task-1-code
#| message: FALSE
# Write your Task 1 code here:
titles <- read_csv("data/titles.csv")
users <- read_csv("data/users.csv")
history <- read_csv("data/history.csv")

# cleaning the titles data
clean_titles <- distinct(titles)
dim(titles)[1] - dim(clean_titles)[1]
# this has removed 10 duplicate rows

# cleaning the history data
clean_history <- distinct(history) %>% 
  mutate("watch_date"= gsub("/","-",watch_date))
dim(history)[1] - dim(clean_history)[1]
# this has removed 311 duplicate rows

# joining the data
cinemetrics <- clean_history %>%
  left_join(clean_titles, by = "movie_id") %>%
  left_join(users, by = "user_id") %>% 
  mutate("watch_date" = as.Date(watch_date))

head(cinemetrics,10)

cinemetrics_messy <- history %>%
  left_join(titles, by = "movie_id") %>%
  left_join(users, by = "user_id")


dim(cinemetrics)[1]
dim(cinemetrics_messy)[1]
```

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation

*Removing duplicates in history was necessary because otherwise duplicate rows would remain after the join, which would skew the analysis of the variables: without cleaning the data first results in a dataset with 22558 rows, whereas with cleaned data we have 19965 rows.*


### Task 2: Missing Values Handling
<!--- DO NOT DELETE THIS LINE - TASK 2 ANCHOR --->

```{r}
#| label: task-2-code
# Write your Task 2 code here:

missing_summary <- cinemetrics %>% 
  is.na() %>% 
  colSums()

missing_summary %>% 
    kable(col.names = c("Variable", "Missing_Count"))

cinemetrics$user_rating_100 %>% 
  median(na.rm = TRUE)

cinemetrics$user_rating_100 %>% 
  replace_na(0) %>% 
  median()
```

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation

*summary() gives a general overview of the data, but may properly account for missing values especially if they are a str or factor. We cannot be sure how it treats missing values and calculates statistics like the median. Therefore it is better to use is.na() to programmatically calculate missing values in our data; this shows us that there are 985 for user_rating_100. We then have freedom over how these are handled, and can decide whther to remove them or assign a specific value depending on context of the data. For example the median after removing the missing values of user_rating_100 is 71, however if we knew that users don't bother rating films they didn't like we could assign missing values as 0, giving us a median of 70. *


### Task 3: Genre Grouping

<!--- DO NOT DELETE THIS LINE - TASK 3 ANCHOR --->

```{r}
#| label: task-3-code
# Write your Task 3 code here:

cinemetrics %>%
  count(genre_tags, sort = TRUE)

unique(cinemetrics$genre_tags)

cinemetrics <- cinemetrics %>% 
  mutate(primary_genre = fct_collapse(factor(genre_tags),
      "Animation" = c("animated", "animation", "animation, kids"),
      "Comedy" = c("comedy, family", "comedy"),
      "Sci-Fi" = c("sci-fi", "Science fiction", "sci-fi, fantasy"),
      "Action" = c("action","action, high-octane"),
      "Romance" = c("romantic comedy", "rom-com", "romantic", "Romance"),
      "Horror" = c("horror", "terror", "scary"),
      "Documentary" = c("documentary, real-world", "Documentary"),
      "Drama" = c("Drama", "drama, emotional", "drama, critically-acclaimed"),
      other_level = "Other"))


unique(cinemetrics$primary_genre)
levels(cinemetrics$primary_genre)

cinemetrics %>%
  count(primary_genre, sort = TRUE) %>%
  kable()


```

```{r}
#| label: task-3-verification
# Write your verification code here:

cinemetrics %>%
  group_by(genre_tags) %>%
  summarise(
    avg_rating = mean(user_rating_100, na.rm = TRUE),  
    movie_count = n()) %>%
  arrange(desc(movie_count)) %>%
  filter(genre_tags %in% c("Romance", "romantic", "rom-com", "romantic comedy")) %>% 
  kable(caption = "Romance vs Rom-Com")

```

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation
When "romantic comedy" movies are collapsed into the "Romance" category, the differences in ratings are lost. The "Romance" tag contains 991 movies with an average rating of 69.83, while the "romantic comedy" category is much smaller with only 145 movies and a significantly lower average rating of 66.35. When these categories are grouped together, the large number of higher-rated "Romance" films dominates the average. As a result, the poorer performance of romantic comedies is masked, creating a statistical distortion as the combined category appears to perform better than the "romantic comedy" films actually do. 

Here, I have decided to group Romantic comedy and Romance together, as romantic comedy is a sub-genre of Romance. Due to both categories having roughly 1500 datapoints, not grouping them together will lead to one of them having to be put into "Other". This would make the other category quite large, and would be primarily made up of one genre which would not make sense and lose us data that we could otherwise make inferences on.


### Task 4: Genre Ratings Table
<!--- DO NOT DELETE THIS LINE - TASK 4 ANCHOR --->

```{r}
#| label: task-4-code
# Write your Task 4 code here:
cinemetrics %>%
  group_by(primary_genre) %>%
  summarise(
    avg_rating = mean(user_rating_100, na.rm = TRUE),  
    movie_count = n()) %>%
  arrange(desc(avg_rating)) %>%
  kable(caption = "Average rating by Genre")

```

```{r}
#| label: task-4-verification
# Write your verification code here:
cinemetrics %>%
  group_by(primary_genre) %>%
  summarise(
    avg_rating = mean(user_rating_100, na.rm = TRUE), 
    avg_budget_USD = mean(budget_usd, na.rm = TRUE),
    movie_count = n()) %>%
  arrange(desc(avg_rating)) %>%
  filter(primary_genre %in% c("Horror", "Action", "Sci-Fi")) %>% 
  kable(caption = "Budget (USD) vs Average rating")

cinemetrics %>% 
  filter(genre_tags %in% c("action", "action, high-octane")) %>% 
  group_by(genre_tags) %>% 
  summarise(avg_rating = mean(user_rating_100, na.rm = TRUE), 
    avg_budget_USD = mean(budget_usd, na.rm = TRUE),
    movie_count = n()) %>% 
  kable(caption = "Budget (USD) vs Average rating for Action subcategories")
```

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation

From our data, there is evidence that larger budget does not necessarily mean higher average rating. Our results show the highest average rating was "Horror" with 79.78, however the average budget for these movies was only approximately $10.2m. We can see that on the other hand, "Sci-Fi" and "Action" both had budgets of over $100m, but despite this they both had average ratings of roughly 69. This shows that larger budget does not necessarily mean users preferred the movies. 

"Action" is a very broad category and may hide information about differences in sub-genres. As there are many different smaller subcategories of action films, by collapsing them into one large category we lose the ability to compare performances of these, and the well-performing ones will be balanced out by the poorly performing ones. In the case of this data, we have only collapsed two sub-genres to make "action" and both of them have ratings differing by only 0.23. Hence our grouping isn't drastically masking the performance of any sub-genres.


### Task 5: Ratings and Engagement Scores
<!--- DO NOT DELETE THIS LINE - TASK 5 ANCHOR --->

```{r}
#| label: task-5-code
# Write your Task 5 code here:

cinemetrics %>% 
  summarise(missing_titles = sum(is.na(movie_title)))
cinemetrics %>% 
  count(movie_title) %>% 
  arrange(n)
cinemetrics %>% 
  count(movie_id) %>% 
  arrange(n)
# we get the same output, so I can be sure there are no formatting errors leading to movies titles being grouped incorrectly

top10_watched_movies <- cinemetrics %>% 
  count(movie_title) %>% 
  arrange(desc(n)) %>% 
  slice_head(n=10)

most_watched_movies <- cinemetrics %>%
  filter(movie_title %in% top10_watched_movies$movie_title) %>% 
  mutate(movie_title = str_wrap(movie_title, width = 10), 
         movie_title = fct_reorder(movie_title, user_rating_100, .desc = TRUE, .na_rm = TRUE)) 
# GenAI used to find str_wrap function, in order to display the movie titles neater in my plot
  

most_watched_movies %>% 
  ggplot(aes(y= user_rating_100))+
  geom_boxplot()+
  facet_wrap(~ movie_title, nrow = 1)+
  labs(y = "Movie Rating (0-100)", title = "Distribution of ratings for top 10 most watched films")+
  theme(axis.ticks.x = element_blank(),
      axis.text.x = element_blank()) # GenAI used to find and understand the theme() function
```

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation

*UP was the movie with the lowest outlier in ratings, at roughly 20/100, meaning one user gave it a very bad rating. However for judging overall audience opinion, we can discount this and look at the IQR. This is relatively small compared to the others in top 10, hence there was less variability in opinions and greater consensus among viewers' ratings.*

```{r}
#| label: task-5-verification
# Write your verification code here:

most_watched_movies %>% 
  group_by(movie_title) %>% 
  summarise(IQR = IQR(user_rating_100, na.rm = TRUE)) %>% 
  arrange(desc(IQR))
```

#### Empirical Verification

*This supports my interpretation, as Up has the second lowest IQR of 19.0, meaning there was less variation in opinions. Inside out is the film with least consensus in opinions with IQR of 22.5, and dune has the greatest consensus in opinions with IQR of 18.0.*


### Task 6: Engagement Score and Subscription Type
<!--- DO NOT DELETE THIS LINE - TASK 6 ANCHOR --->

```{r}
#| label: task-6-code
# Write your Task 6 code here:

cinemetrics %>% 
  filter(!is.na(subscription_type)) %>% 
  group_by(subscription_type) %>% 
  summarise(avg_engagement = mean(engagement_score, na.rm = TRUE)) %>% 
  arrange(desc(avg_engagement)) %>% 
  kable()

age_quantiles <- quantile(cinemetrics$age, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)

# gen AI was used to understand the probs argument in quantile()

age_quantiles %>% 
  kable(col.names = c("Quantile", "Value"), caption = "User age quantiles")
  
cinemetrics <- cinemetrics %>% 
  mutate(age_tier = case_when(age <= age_quantiles[1] ~ "1",
                              age <= age_quantiles[2] ~ "2",
                              age <= age_quantiles[3] ~ "3",
                              age > age_quantiles[3]  ~ "4"))

avg_engagement <- cinemetrics %>% 
  filter(!is.na(subscription_type), !is.na(age_tier)) %>% 
  group_by(age_tier, subscription_type) %>% 
  summarise(
    avg_engagement = mean(engagement_score, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  arrange(age_tier) 

avg_engagement %>% 
  pivot_wider(names_from = subscription_type, values_from = avg_engagement) %>% 
  kable(caption = "Average Engagement")

```

```{r}
#| label: task-6-verification
# Write your verification code here:

age_subtype_count <- cinemetrics %>% 
  filter(!is.na(subscription_type), !is.na(age_tier)) %>% 
  group_by(age_tier, subscription_type) %>% 
  summarise(total = n(), .groups = "drop") %>% 
  pivot_wider(names_from = subscription_type,
              values_from = total)
  
age_subtype_count %>% 
  kable(caption = "No. users per subscription type by age quartile")
```

#### Empirical Verification

*For users 25 or younger, 91.0% use the basic subscription type. As age increases, proportion of users with basic decreases. Only 13.3% of users over 45 have a basic subscription,the rest have premium.*

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation

*Our 2D analysis shows that engagement score for premium users (56.5) is higher than that of basic users (44.0). However in our analysis split by age category, basic users have higher engagement than premium for all age categories. This is an example of "Simpson's Paradox": the change in observed trend is due to the differences in group sizes and distribution of users. For instance, in this scenario premium users are more highly concentrated in higher age categories (as seen in verification), where engagement levels are higher generally, hence the average engagement is weighted towards premium. On the other end, 91.0% of users 25 or younger have basic subscription, but the engagement in age tier 1 is significantly lower in general (28.9 for basic and 12.6 for premium) hence engagement for basic users is a lot lower, despite being higher in every individual age tier.*


---

## Part B: Text Data Analysis

### Task 7: Tokenization & Word Frequency
<!--- DO NOT DELETE THIS LINE - TASK 7 ANCHOR --->

```{r}
#| label: task-7-code
# Write your Task 7 code here:

# create an artificial review id category so that we can distinguish between reviews
cinemetrics <- cinemetrics %>%
  mutate(review_id = row_number())

words <- cinemetrics %>% 
  unnest_tokens(word, review_snippet) %>% 
  anti_join(stop_words, join_by(word))

top_10_words <- words %>%
  count(word, sort = TRUE) %>% 
  slice_head(n=10)

top_10_words %>% 
  ggplot(aes(x = fct_reorder(word,-n), y = n))+
  geom_col(fill = "firebrick")+
  labs(title = "Top 10 Most Frequent Words in Movie Reviews", x = "Word", y = "Frequency")
  

```

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation

*Even after removing stop words, the top two words are "film" (5674 counts) and "watch" (5317 counts) which are unhelpful for analysis as they tell us nothing about the the opinion of the user. They are generic and we cannot infer any useful information from them, such as whether the user liked the film.*


### Task 8: Sentiment Analysis
<!--- DO NOT DELETE THIS LINE - TASK 8 ANCHOR --->

```{r}
#| label: task-8-code
# Write your Task 8 code here:

bing_lexicon <- get_sentiments("bing")

review_sentiment <- words %>% 
  inner_join(bing_lexicon, by = "word") 

sentiment_count <- review_sentiment %>% 
    count(review_id, sentiment) %>% 
    pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>% 
    mutate(net_sentiment = positive - negative)

head(sentiment_count,50)

cinemetrics <- cinemetrics %>% 
  left_join(sentiment_count, join_by(review_id))

sentiment_v_rating <- cinemetrics %>% 
  group_by(primary_genre) %>% 
  summarise(
    avg_net_sentiment = mean(net_sentiment, na.rm = TRUE),
    avg_user_rating = mean(user_rating_100, na.rm = TRUE))

sentiment_v_rating %>% 
  kable(caption = "Overall sentiment & rating of film genres")

plot_sentiment_rating <- sentiment_v_rating %>% 
  ggplot(aes(x = avg_net_sentiment, y = avg_user_rating)) +
  geom_point() +
  labs(
    x = "Average Sentiment",
    y = "Average User Rating",
    title = "Average Sentiment vs Average Rating by Genre"
  ) +
  theme_minimal()+
  theme(plot.title = element_text(size = 10))

plot_remove_outlier <- sentiment_v_rating %>% 
  slice(-6) %>% 
  ggplot(aes(x = avg_net_sentiment, y = avg_user_rating)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE)+
  labs(
    x = "Average Sentiment",
    y = "",
    title = "Average Sentiment vs Average Rating by Genre (removing outliers)"
  ) +
  theme_minimal()+
  theme(plot.title = element_text(size = 7))

plot_sentiment_rating + plot_remove_outlier
```

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation

*The outlier on this dataset is "Horror" as it has significantly higher rating of 79.8, whilst having a sentiment of -1.14. This is unsurprising, as many adjectives that would be used to describe horror films are typically classed as negative, such as "scary", "terrifying" etc. However when in this context, these are actually positive remarks on the movie, as that is what people want from a horror movie. If we remove this outlier and look at the remaining points, we see they have slight positive correlation. This is what we would expect to see, as when the sentiment of reviews grows more positive the ratings are also likely to be higher, as the users clearly liked the movie more.*


### Task 9: Advanced Text Mining (TF-IDF)
<!--- DO NOT DELETE THIS LINE - TASK 9 ANCHOR --->

```{r}
#| label: task-9-code
# Write your Task 9 code here: 

genre_words <- words %>% 
  count(primary_genre, word, sort = TRUE)

genre_tf_idf <- genre_words  %>% 
  bind_tf_idf(term = word, document = primary_genre, n = n)

top_tf_idf_by_genre <- genre_tf_idf %>% 
  group_by(primary_genre) %>% 
  slice_max(order_by = tf_idf,n = 1) %>% 
  arrange(primary_genre, desc(tf_idf))


# following code was written for the written interpretation section

# 1
genre_tf_idf %>% 
  filter(primary_genre == "Romance", word == "film") %>% 
  select(-n)

# 3
genre_tf_idf %>% 
  filter(primary_genre == "Romance") %>% 
  arrange(desc(tf_idf)) %>% 
  slice_head(n=3)

genre_tf_idf %>% 
  filter(primary_genre == "Romance") %>% 
  arrange(desc(n)) %>% 
  slice_head(n=3)
```

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation

1. 0
2. *TF-IDF is calculated by multiplying the term frequency in a specific genre by log(number of genres/number of genres the term appears in). The term “film” appears in reviews for every genre, so the IDF becomes log(8/8)=0, and hence a TF-IDF score of 0. This makes sense because IDF weights terms by how unique they are across genres, and “film” is not unique to any specific genre.*
3. *The most frequent words in Romance are "film" (697), "watch" (680), "average" (359). These are generic words that don't tell us much about Romance as a genre specifically. The top 3 highest TF-IDF words are "watch" (0.014), "average" (0.007), "fairly" (0.007). In this scenario, due only having 8 genres, and the language used across reviews for any film is quite similar, the IDF component does not distinguish terms as effectively. It does help a little bit though, by removing "film" which is the most generic term. Instead we see "fairly" which does give us a little bit more insight into the general strength of opinions about Romance films.*



### Task 10: Executive Recommendation
<!--- DO NOT DELETE THIS LINE - TASK 10 ANCHOR --->

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation

*Replace this text with your one-paragraph executive recommendation (max 150 words), citing at least three specific findings from earlier tasks and acknowledging one limitation.*


---
<!--- WORD COUNT ANCHOR --->
```{r}
#| label: word-count
#| echo: false
#| message: false
#| warning: false

count_prose_words <- function(file_path) {
  # Read the raw lines of the file
  lines <- readLines(file_path, warn = FALSE)
  
  # 1. Strip YAML Header (Lines between the first two '---')
  yaml_bounds <- grep("^---$", lines)
  if (length(yaml_bounds) >= 2) {
    lines <- c(lines[1:(yaml_bounds[1]-1)], lines[(yaml_bounds[2]+1):length(lines)])
  }
  
  # 2. Strip Code Chunks (Lines between pairs of '```')
  chunk_bounds <- grep("^```", lines)
  if (length(chunk_bounds) %% 2 == 0 && length(chunk_bounds) > 0) {
    # Create a vector of line indices to remove
    lines_to_remove <- integer(0)
    for (i in seq(1, length(chunk_bounds), by = 2)) {
      lines_to_remove <- c(lines_to_remove, chunk_bounds[i]:chunk_bounds[i+1])
    }
    lines <- lines[-lines_to_remove]
  }
  
  # 3. Strip Markdown Headers (Lines starting with #)
  lines <- lines[!grepl("^#+", trimws(lines))]
  
  # 4. Strip HTML Comments (Lines starting with <!--)
  lines <- lines[!grepl("^<!--", trimws(lines))]
  
  # 5. Combine what's left into a single string
  text_only <- paste(lines, collapse = " ")
  
  # 6. Remove common markdown characters that might be counted as words
  text_only <- gsub("[*>`-]", " ", text_only)
  
  # 7. Count the remaining words
  words <- unlist(strsplit(trimws(text_only), "\\s+"))
  words <- words[words != ""]
  
  return(length(words))
}

# Run the function on the current file
true_word_count <- count_prose_words(knitr::current_input())

# Print the final result formatted nicely
cat(sprintf("**Prose Word Count:** %d words", true_word_count))
```
