---
title: "Coursework 1: MA22019, Introduction to Data Science"
author: "Ewan Brett"
format: gfm
---

<!--- DO NOT DELETE THIS LINE --->
<!--- METADATA ANCHOR: MA22019-CW1-2026 --->

Please read the `coursework_01_instructions.pdf` for the full context, data dictionary, and detailed task requirements before starting.



```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(tidytext)
library(knitr)
library(patchwork)
```

## Part A: Data Wrangling & EDA 

### Task 1: Data Joining
<!--- DO NOT DELETE THIS LINE - TASK 1 ANCHOR --->

```{r}
#| label: task-1-code
#| message: FALSE
# Write your Task 1 code here:
titles <- read_csv("data/titles.csv")
users <- read_csv("data/users.csv")
history <- read_csv("data/history.csv")

# cleaning the titles data
clean_titles <- distinct(titles)
dim(titles)[1] - dim(clean_titles)[1]
# this has removed 10 duplicate rows

# cleaning the history data
clean_history <- distinct(history) %>% 
  mutate("watch_date"= gsub("/","-",watch_date))
dim(history)[1] - dim(clean_history)[1]
# this has removed 311 duplicate rows

# joining the data
cinemetrics <- clean_history %>%
  left_join(clean_titles, by = "movie_id") %>%
  left_join(users, by = "user_id") %>% 
  mutate("watch_date" = as.Date(watch_date))

head(cinemetrics,10)

cinemetrics_messy <- history %>%
  left_join(titles, by = "movie_id") %>%
  left_join(users, by = "user_id")


dim(cinemetrics)[1]
dim(cinemetrics_messy)[1]
```

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation

*Removing duplicates in history was necessary because otherwise duplicate rows would remain after the join, which would skew the analysis of the variables. Performing the join without cleaning the data first results in a dataset with 22558 rows, whereas with cleaned data we have 19965 rows.*


### Task 2: Missing Values Handling
<!--- DO NOT DELETE THIS LINE - TASK 2 ANCHOR --->

```{r}
#| label: task-2-code
# Write your Task 2 code here:

missing_summary <- cinemetrics %>% 
  is.na() %>% 
  colSums()

missing_summary %>% 
    kable(col.names = c("Variable", "Missing_Count"))

cinemetrics$user_rating_100 %>% 
  median(na.rm = TRUE)

cinemetrics$user_rating_100 %>% 
  replace_na(0) %>% 
  median()
```

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation

*summary() gives a general overview of the data, but may not account for missing values, especially if the variable is a str or a factor. We cannot be sure how it treats missing values, and calculates statistics like the median. Hence it is much more reliable to use is.na() to programmatically calculate the missing values in our data, and this shows us that there are 985 for user_rating_100. That way we have the freedom over how these are handled, e.g. whether you remove missing values, or assign them a specific value, this can be decided based on context of the data which summary() wouldn't handle as effectively. For example the median after removing the missing values of user_rating_100 is 71, however if we knew that users don't bother rating films they didn't like, then it would make more sense to assign missing values as 0, giving us a median of 70. *


### Task 3: Genre Grouping

<!--- DO NOT DELETE THIS LINE - TASK 3 ANCHOR --->

```{r}
#| label: task-3-code
# Write your Task 3 code here:

cinemetrics %>%
  count(genre_tags, sort = TRUE)

unique(cinemetrics$genre_tags)

cinemetrics <- cinemetrics %>% 
  mutate(primary_genre = fct_collapse(factor(genre_tags),
      "Animation" = c("animated", "animation", "animation, kids"),
      "Comedy" = c("comedy, family", "comedy"),
      "Sci-Fi" = c("sci-fi", "Science fiction", "sci-fi, fantasy"),
      "Action" = c("action","action, high-octane"),
      "Romance" = c("romantic comedy", "rom-com", "romantic", "Romance"),
      "Horror" = c("horror", "terror", "scary"),
      "Documentary" = c("documentary, real-world", "Documentary"),
      "Drama" = c("Drama", "drama, emotional", "drama, critically-acclaimed"),
      other_level = "Other"))


unique(cinemetrics$primary_genre)
levels(cinemetrics$primary_genre)

cinemetrics %>%
  count(primary_genre, sort = TRUE) %>%
  kable()


```

```{r}
#| label: task-3-verification
# Write your verification code here:

cinemetrics %>%
  group_by(genre_tags) %>%
  summarise(
    avg_rating = mean(user_rating_100, na.rm = TRUE),  
    movie_count = n()) %>%
  arrange(desc(movie_count)) %>%
  filter(genre_tags %in% c("Romance", "romantic", "rom-com", "romantic comedy")) %>% 
  kable(caption = "Romance vs Rom-Com")

```

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation

NEED TO CHANGE THIS:


*When "romantic comedy" movies are collapsed into the "Romance" category, the differences in ratings are lost. The "Romance" tag contains 991 movies with an average rating of 69.83, while the "romantic comedy" category is much smaller with only 145 movies and a significantly lower average rating of 66.35. When these categories are grouped together, the large number of higher-rated "Romance" films dominates the average. As a result, the poorer performance of romantic comedies is masked, creating a statistical distortion as the combined category appears to perform better than the "romantic comedy" films actually do.


This table shows us the original "genre_tags" has inconsistent labeling which makes it unsuitable to perform proper analysis. We see there are larger categories such as "rom-com" (1009 movies) and "romance" (991 movies), however we also have smaller categories representing the same group but differently names, such as "romantic" (150) and "romantic comedy" (145). Clearly "rom-com" and "romantic comedy" should be put in the same group and "romance" and "romantic" should also be the same group. However we then encounter a potential error which could effect our analysis, as "romantic" is just an adjective, and potentially there could be Rom-Coms in that category, which would be mislabeled if we kept Rom-Com and Romance as seperate categories. Further, I found that when we keep them as seperate categories, and group by the most frequent five categories, "Other" becomes the largest category with 4225 entries. It does not make sense to have such a large "Other" category when these can be grouped into bigger categories with similar properties. When we group them together, "Other" has only 2295 entries, hence we have more insightful categories. Hence, I have decided grouping Rom-coms and Romances as one category will be the best for our analysis, and allow for better interpretability.*


### Task 4: Genre Ratings Table
<!--- DO NOT DELETE THIS LINE - TASK 4 ANCHOR --->

```{r}
#| label: task-4-code
# Write your Task 4 code here:
cinemetrics %>%
  group_by(primary_genre) %>%
  summarise(
    avg_rating = mean(user_rating_100, na.rm = TRUE),  
    movie_count = n()) %>%
  arrange(desc(avg_rating)) %>%
  kable(caption = "Average rating by Genre")

```

```{r}
#| label: task-4-verification
# Write your verification code here:
cinemetrics %>%
  group_by(primary_genre) %>%
  summarise(
    avg_rating = mean(user_rating_100, na.rm = TRUE), 
    avg_budget_USD = mean(budget_usd, na.rm = TRUE),
    movie_count = n()) %>%
  arrange(desc(avg_rating)) %>%
  filter(primary_genre %in% c("Horror", "Action", "Sci-Fi")) %>% 
  kable(caption = "Budget (USD) vs Average rating")

cinemetrics %>% 
  filter(genre_tags %in% c("action", "action, high-octane")) %>% 
  group_by(genre_tags) %>% 
  summarise(avg_rating = mean(user_rating_100, na.rm = TRUE), 
    avg_budget_USD = mean(budget_usd, na.rm = TRUE),
    movie_count = n()) %>% 
  kable(caption = "Budget (USD) vs Average rating for Action subcategories")
```

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation

*From our data, there is evidence that larger budget does not necessarily mean higher average rating. Our results show the highest average rating was "Horror" with 79.78, however the average budget for these movies was only approximately $10.2m. We can see that on the other hand, "Sci-Fi" and "Action" both had budgets of over $100m, but despite this they both had average ratings of roughly 69. This shows that larger budget does not necessarily mean users preferred the movies. 

"Action" is a very broad category and may hide valuable information about differences in sub-genres. As there are many different smaller subcategories of action films, by collapsing them into one large category we lose the ability to compare performances of these sub-genres, and the well-performing ones will be balanced out by the poorly performing ones. In the case of this data, we have only collapsed two sub-genres to make "action" and both of them have ratings of roughly 69, differing by only 0.3. Hence our grouping isn't masking the performance of any sub-genres in this case. 


Sci-Fi has the largest budget (139588009USD), however it has the second lowest average rating. We see here that Horror and Action are not part of the top 5 categories, and therefore come under "Other". The average budget for "Other" (58846173USD) is roughly half of the Sci-Fi budget, however the average ratings (74.35) are much higher. Hence larger budget does not guarantee higher average rating. We can look at the unlumped categories, which further reinforces this as Horror has by far the best ratings (79.78), however has the smallest budget (10181094USD), and Action has the lowest rating (69.22) and a much larger budget (115543533USD). A very broad category like "Action" can mean that performance of many different sub-categories balance out, so we cannot really make good inferences from the data. For example, more niche categories which don't perform well won't show in our inferences as they will be averaged out by more successful subcategories within "Action".*


### Task 5: Ratings and Engagement Scores
<!--- DO NOT DELETE THIS LINE - TASK 5 ANCHOR --->

```{r}
#| label: task-5-code
# Write your Task 5 code here:

cinemetrics %>% 
  summarise(missing_titles = sum(is.na(movie_title)))
cinemetrics %>% 
  count(movie_title) %>% 
  arrange(n)
cinemetrics %>% 
  count(movie_id) %>% 
  arrange(n)
# we get the same output, so I can be sure there are no formatting errors leading to movies titles being grouped incorrectly

top10_watched_movies <- cinemetrics %>% 
  count(movie_title) %>% 
  arrange(desc(n)) %>% 
  slice_head(n=10)

most_watched_movies <- cinemetrics %>%
  filter(movie_title %in% top10_watched_movies$movie_title) %>% 
  mutate(movie_title = str_wrap(movie_title, width = 10), 
         movie_title = fct_reorder(movie_title, user_rating_100, .desc = TRUE, .na_rm = TRUE)) 
# GenAI used to find str_wrap function, in order to display the movie titles neater in my plot
  

most_watched_movies %>% 
  ggplot(aes(y= user_rating_100))+
  geom_boxplot()+
  facet_wrap(~ movie_title, nrow = 1)+
  labs(y = "Movie Rating (0-100)", title = "Distribution of ratings for top 10 most watched films")+
  theme(axis.ticks.x = element_blank(),
      axis.text.x = element_blank()) # GenAI used to find and understand the theme() function
```

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation

*UP was the movie with the lowest outlier in ratings, at roughly 20/100, meaning one user gave it a very bad rating. However for judging overall audience opinion, we can discount this and look at the IQR. This is relatively small compared to the others in top 10, hence there was less variability in opinions and greater consensus among viewers' ratings.*

```{r}
#| label: task-5-verification
# Write your verification code here:

most_watched_movies %>% 
  group_by(movie_title) %>% 
  summarise(IQR = IQR(user_rating_100, na.rm = TRUE)) %>% 
  arrange(desc(IQR))
```

#### Empirical Verification

*This supports my interpretation, as Up has the second lowest IQR of 19.0, meaning there was less variation in opinions. Inside out is the film with least consensus in opinions with IQR of 22.5.*


### Task 6: Engagement Score and Subscription Type
<!--- DO NOT DELETE THIS LINE - TASK 6 ANCHOR --->

```{r}
#| label: task-6-code
# Write your Task 6 code here:

cinemetrics %>% 
  filter(!is.na(subscription_type)) %>% 
  group_by(subscription_type) %>% 
  summarise(avg_engagement = mean(engagement_score, na.rm = TRUE)) %>% 
  arrange(desc(avg_engagement)) %>% 
  kable()

age_quantiles <- quantile(cinemetrics$age, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)

# gen AI was used to understand the probs argument in quantile()

age_quantiles %>% 
  kable(col.names = c("Quantile", "Value"), caption = "User age quantiles")
  
cinemetrics <- cinemetrics %>% 
  mutate(age_tier = case_when(age <= age_quantiles[1] ~ "1",
                              age <= age_quantiles[2] ~ "2",
                              age <= age_quantiles[3] ~ "3",
                              age > age_quantiles[3]  ~ "4"))

avg_engagement <- cinemetrics %>% 
  filter(!is.na(subscription_type), !is.na(age_tier)) %>% 
  group_by(age_tier, subscription_type) %>% 
  summarise(
    avg_engagement = mean(engagement_score, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  arrange(age_tier) 

avg_engagement %>% 
  pivot_wider(names_from = subscription_type, values_from = avg_engagement) %>% 
  kable(caption = "Average Engagement")

```

```{r}
#| label: task-6-verification
# Write your verification code here:

age_subtype_count <- cinemetrics %>% 
  filter(!is.na(subscription_type), !is.na(age_tier)) %>% 
  group_by(age_tier, subscription_type) %>% 
  summarise(total = n(), .groups = "drop") %>% 
  pivot_wider(names_from = subscription_type,
              values_from = total)
  
age_subtype_count %>% 
  kable(caption = "No. users per subscription type by age quartile")
```

#### Empirical Verification

*For users 25 or younger, 91.0% use the basic subscription type. As age increases, proportion of users with basic decreases. Only 13.3% users over 45 have a basic subscription.*

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation

*Our 2D analysis shows that engagement score for premium users (56.5) is higher than that of basic users (44.0). However in our analysis split by age category, basic users have higher engagement than premium for all age categories. This change in observed trend is due to the differences in group sizes and distribution of users. For instance, in this scenario premium users are more highly concentrated in higher age categories (as seen in verification), where engagement levels are higher generally, hence the average engagement is weighted towards premium. *


---

## Part B: Text Data Analysis

### Task 7: Tokenization & Word Frequency
<!--- DO NOT DELETE THIS LINE - TASK 7 ANCHOR --->

```{r}
#| label: task-7-code
# Write your Task 7 code here:

# create an artificial review id category so that we can distinguish between reviews
cinemetrics <- cinemetrics %>%
  mutate(review_id = row_number())

words <- cinemetrics %>% 
  unnest_tokens(word, review_snippet) %>% 
  anti_join(stop_words, join_by(word))

top_10_words <- words %>%
  count(word, sort = TRUE) %>% 
  slice_head(n=10)

top_10_words %>% 
  ggplot(aes(x = fct_reorder(word,-n), y = n))+
  geom_col(fill = "firebrick")+
  labs(title = "Top 10 Most Frequent Words in Movie Reviews", x = "Word", y = "Frequency")
  

```

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation

*The top two words are "film" (5674 counts) and "watch" (5317 counts) which are unhelpful for analysis as they tell us nothing about the the opinion of the user. They are generic and we cannot infer any information from them, such as whether the user liked the film.*


### Task 8: Sentiment Analysis
<!--- DO NOT DELETE THIS LINE - TASK 8 ANCHOR --->

```{r}
#| label: task-8-code
# Write your Task 8 code here:

bing_lexicon <- get_sentiments("bing")

review_sentiment <- words %>% 
  inner_join(bing_lexicon, by = "word") 

sentiment_count <- review_sentiment %>% 
    count(review_id, sentiment) %>% 
    pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>% 
    mutate(net_sentiment = positive - negative)

head(sentiment_count,50)

cinemetrics <- cinemetrics %>% 
  left_join(sentiment_count, join_by(review_id))

sentiment_v_rating <- cinemetrics %>% 
  group_by(primary_genre) %>% 
  summarise(
    avg_net_sentiment = mean(net_sentiment, na.rm = TRUE),
    avg_user_rating = mean(user_rating_100, na.rm = TRUE))

sentiment_v_rating %>% 
  kable(caption = "Overall sentiment & rating of film genres")

plot_sentiment_rating <- sentiment_v_rating %>% 
  ggplot(aes(x = avg_net_sentiment, y = avg_user_rating)) +
  geom_point() +
  labs(
    x = "Average Sentiment",
    y = "Average User Rating",
    title = "Average Sentiment vs Average Rating by Genre"
  ) +
  theme_minimal()+
  theme(plot.title = element_text(size = 10))

plot_remove_outlier <- sentiment_v_rating %>% 
  slice(-6) %>% 
  ggplot(aes(x = avg_net_sentiment, y = avg_user_rating)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE)+
  labs(
    x = "Average Sentiment",
    y = "",
    title = "Average Sentiment vs Average Rating by Genre (removing outliers)"
  ) +
  theme_minimal()+
  theme(plot.title = element_text(size = 7))

plot_sentiment_rating + plot_remove_outlier
```

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation

*Replace this text with your explanation of the scatterplot outlier.*


### Task 9: Advanced Text Mining (TF-IDF)
<!--- DO NOT DELETE THIS LINE - TASK 9 ANCHOR --->

```{r}
#| label: task-9-code
# Write your Task 9 code here: 

genre_words <- words %>% 
  count(primary_genre, word, sort = TRUE)

genre_tf_idf <- genre_words  %>% 
  bind_tf_idf(term = word, document = primary_genre, n = n)

top_tf_idf_by_genre <- genre_tf_idf %>% 
  group_by(primary_genre) %>% 
  slice_max(order_by = tf_idf,n = 1) %>% 
  arrange(primary_genre, desc(tf_idf))

```

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation

1. *Replace this text with the TF-IDF score for the word "film".*
2. *Replace this text with your mathematical explanation using the IDF equation from the lecture.*
3. *Replace this text with your comparison of TF-IDF vs frequency.*



### Task 10: Executive Recommendation
<!--- DO NOT DELETE THIS LINE - TASK 10 ANCHOR --->

<!--- WRITE YOUR WRITTEN ANSWER BELOW THIS LINE --->

#### Written Interpretation

*Replace this text with your one-paragraph executive recommendation (max 150 words), citing at least three specific findings from earlier tasks and acknowledging one limitation.*


---
<!--- WORD COUNT ANCHOR --->
```{r}
#| label: word-count
#| echo: false
#| message: false
#| warning: false

count_prose_words <- function(file_path) {
  # Read the raw lines of the file
  lines <- readLines(file_path, warn = FALSE)
  
  # 1. Strip YAML Header (Lines between the first two '---')
  yaml_bounds <- grep("^---$", lines)
  if (length(yaml_bounds) >= 2) {
    lines <- c(lines[1:(yaml_bounds[1]-1)], lines[(yaml_bounds[2]+1):length(lines)])
  }
  
  # 2. Strip Code Chunks (Lines between pairs of '```')
  chunk_bounds <- grep("^```", lines)
  if (length(chunk_bounds) %% 2 == 0 && length(chunk_bounds) > 0) {
    # Create a vector of line indices to remove
    lines_to_remove <- integer(0)
    for (i in seq(1, length(chunk_bounds), by = 2)) {
      lines_to_remove <- c(lines_to_remove, chunk_bounds[i]:chunk_bounds[i+1])
    }
    lines <- lines[-lines_to_remove]
  }
  
  # 3. Strip Markdown Headers (Lines starting with #)
  lines <- lines[!grepl("^#+", trimws(lines))]
  
  # 4. Strip HTML Comments (Lines starting with <!--)
  lines <- lines[!grepl("^<!--", trimws(lines))]
  
  # 5. Combine what's left into a single string
  text_only <- paste(lines, collapse = " ")
  
  # 6. Remove common markdown characters that might be counted as words
  text_only <- gsub("[*>`-]", " ", text_only)
  
  # 7. Count the remaining words
  words <- unlist(strsplit(trimws(text_only), "\\s+"))
  words <- words[words != ""]
  
  return(length(words))
}

# Run the function on the current file
true_word_count <- count_prose_words(knitr::current_input())

# Print the final result formatted nicely
cat(sprintf("**Prose Word Count:** %d words", true_word_count))
```
